---
title: "Merge .h5 files by Sample/HTO Barcode"
author: 
 - Lucas Graybuck
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    self_contained: true
params:
  in_dir: NULL
  out_dir: NULL
---

## Contents

#### [Data Processing](#data_processing)
- [Session Preparation](#session_preparation)
- [Identify Inputs](#find_inputs)
- [Identify Hashes](#find_hashes)
- [Merge by Hash](#merge_data)

#### [QC Metrics and Plots](#qc_metrics)
- [Pool-Based QC](#pool_qc)
- [Well-Based QC](#well_qc)
  - [Counts, Reads, Genes, and UMIs](#well_overview)
  - [HTO Category by Well](#well_category_counts)
  - [HTO Barcodes by Well](#well_barcode_counts)
- [HTO-Based QC](#hash_qc)
  - [Counts](#hash_counts)
  - [Reads](#hash_reads)
  - [UMIs](#hash_umis)
  - [Genes](#hash_genes)
- [Write QC JSON](#json_out)

#### [Session Info](#session_info)

<a id="data_processing"></a>

## Data Processing

<a id="session_preparation"></a>

### Session Preparation

#### Load libraries:
```{r Load Libraries}
start_time <- Sys.time()

quiet_library <- function(...) {
  suppressPackageStartupMessages(library(...))
}
quiet_library(rhdf5)
quiet_library(BarMixer)
quiet_library(Matrix)
quiet_library(ggplot2)
quiet_library(cowplot)
```

Declaring start
```{r Declare start}
stm("Merging .h5 files")
stm(paste0("Using BarMixer v", installed.packages()["BarMixer","Version"]))
```

#### Argument parsing
```{r Parse arguments}
in_dir <- params$in_dir
out_dir <- params$out_dir

stm(paste0("IN  H5 directory    : ", in_dir))
stm(paste0("OUT Directory       : ", out_dir))
```

```{r Print Arguments}
print(c(
  paste0("IN  H5 directory    : ", in_dir),
  paste0("OUT Directory       : ", out_dir)
))
```

[Return to Contents](#contents)

<a id="find_inputs"></a>

#### Locate input files
```{r Locate inputs}
in_file_names <- list.files(in_dir, 
                            pattern = "_.+h5$")
# Check for per-well directories
if(length(in_file_names) == 0) {
  in_file_dirs <- list.dirs(in_dir, 
                            recursive = FALSE)
  in_file_names <- unlist(lapply(in_file_dirs,
                                 list.files, 
                                 pattern = "_.+h5$"))
  in_file_paths <- unlist(lapply(in_file_dirs,
                                 list.files,
                                 pattern = "_.+h5$", 
                                 full.names = TRUE))
} else {
  in_file_paths <- list.files(in_dir, 
                              pattern = "_.+h5$", 
                              full.names = TRUE)
}

stm(paste0("Found ", length(in_file_paths), " .h5 files for processing"))
print(paste0("Found ", length(in_file_paths), " .h5 files for processing"))
```

[Return to Contents](#contents)

<a id="find_hashes"></a>

#### Identify hashes in in_dir
```{r Identify hashes}
in_file_hashes <- sub(".h5$","",sub(".+_","",in_file_names))
unique_hashes <- unique(in_file_hashes)

stm(paste0("Identified ", length(unique_hashes), " hash patterns: ", paste(unique_hashes, collapse = ", ")))
```

[Return to Contents](#contents)

<a id="merge_data"></a>

#### Merge files based on hashes
```{r Load Count Matrix}
meta_list <- list()
for(i in seq_along(unique_hashes)) {
  current_hash <- unique_hashes[i]

  stm(paste0("Merging for hash ", current_hash))
  current_files <- in_file_paths[in_file_hashes == current_hash]
  for(f in current_files) {
    if(file.exists(f)) {
      stm(paste0("IN  H5 file         : ", f))
    } else {
      stm(paste0("MISSING  H5 file    : ", f))
      current_files <- current_files[current_files != f]
    }
  }
  
  if(length(current_files) > 1) {
    h5_ll <- lapply(current_files,
                    h5dump)
    
    h5_list <- reduce_h5_list(h5_ll)
    
  } else {
    h5_list <- h5dump(current_files)
  }
  
  meta_list[[current_hash]] <- h5_list_cell_metadata(h5_list)
  
  current_pool <- meta_list[[current_hash]]$pool_id[1]
  
  if(current_hash == "multiplet") {
    out_file <- file.path(out_dir, 
                          paste0(current_pool, "_",
                                 current_hash, ".h5"))
    t_out_file <- file.path(out_dir, 
                            paste0("transposed_",
                                   current_pool, "_",
                                   current_hash, ".h5"))
  } else {
    current_pbmc_id <- meta_list[[current_hash]]$pbmc_sample_id[1]
    out_file <- file.path(out_dir, 
                          paste0(current_pool, "_",
                                 current_pbmc_id, ".h5"))
    t_out_file <- file.path(out_dir, 
                            paste0("transposed_",
                                   current_pool, "_",
                                   current_pbmc_id, ".h5"))
  }
  
  stm(paste0("Writing HDF5 to     : ", out_file))
  
  write_h5_list(h5_list,
                out_file,
                overwrite = TRUE)
}
```

[Return to Contents](#contents)

<a id="qc_metrics"></a>

## QC Metrics

#### Metrics Setup
```{r Declare Metrics and Plotting}
stm("Generating QC Metrics and Plots for Report")
all_meta <- do.call("rbind", meta_list)
all_meta <- as.data.table(all_meta)
if("pbmc_sample_id" %in% names(all_meta)) {
  all_meta$plot_barcode <- paste0(all_meta$hto_barcode,
                                  "\n",
                                  all_meta$pbmc_sample_id)
  
  hto_barcode_to_pbmc_id <- data.frame(hto_barcode = unique(all_meta$hto_barcode),
                                       pbmc_sample_id = unique(all_meta$pbmc_sample_id))
} else {
  all_meta$pbmc_sample_id <- "No PBMC ID"
  all_meta$plot_barcode <- all_meta$hto_barcode
  
  hto_barcode_to_pbmc_id <- data.frame(pbmc_sample_id = "No PBMC ID",
                                       hto_barcode = unique(all_meta$hto_barcode))
}

singlet_meta <- all_meta[hto_category == "singlet"]
hto_barcode_to_pbmc_id <- as.data.table(hto_barcode_to_pbmc_id)

qc_list <- list(report_type = "merge_h5_by_hash",
                report_datetime = as.character(start_time),
                report_uuid = ids::uuid(use_time = TRUE),
                package = "BarMixer",
                package_version = sessionInfo()$otherPkgs$BarMixer$Version,
                pool_id = current_pool)

out_json <- file.path(in_dir, paste0(current_pool, "_merge_h5_metrics.json"))
stm(paste0("OUT JSON            : ", out_json))
print(paste0("OUT JSON            : ", out_json))
```

[Return to Contents](#contents)

<a id="pool_qc"></a>

### Pool-Based QC

#### Counts, Median Reads, UMIs, Genes across all wells
```{r}
overview_metrics <- all_meta[,.(n_cells = nrow(.SD),
                                median_reads = median(n_reads),
                                median_umis = median(n_umis),
                                median_genes = median(n_genes))]

qc_list$pool_stats <- as.list(overview_metrics)

qc_table(overview_metrics)
```

[Return to Contents](#contents)

<a id="pool_category_metrics"></a>

#### Counts, Median Reads, UMIs, Genes by HTO Category
```{r}
hto_category_metrics <- all_meta[,.(n_cells = nrow(.SD),
                                    median_reads = as.numeric(median(n_reads)),
                                    median_umis = as.numeric(median(n_umis)),
                                    median_genes = as.numeric(median(n_genes))),
                                 by = hto_category]

qc_list$hto_category_stats <- lapply(1:nrow(hto_category_metrics),
                                      function(x) {
                                        list(n_cells = hto_category_metrics$n_cells[x],
                                             median_reads = hto_category_metrics$median_reads[x],
                                             median_umis = hto_category_metrics$median_umis[x],
                                             median_genes = hto_category_metrics$median_genes[x])
                                      })
names(qc_list$hto_category_stats) <- hto_category_metrics$hto_category

qc_table(hto_category_metrics)
```

[Return to Contents](#contents)

<a id="well_qc"></a>

### Well-Based QC

<a id="well_overview"></a>

#### Counts, Median Reads, UMIs, Genes per well
```{r}
well_metrics <- all_meta[,.(n_cells = nrow(.SD),
                            median_reads = as.numeric(median(n_reads)),
                            median_umis = as.numeric(median(n_umis)),
                            median_genes = as.numeric(median(n_genes))),
                         by = well_id]

for(well in well_metrics$well_id) {
  qc_list$well_stats[[well]] <- list(n_cells = well_metrics$n_cells[well_metrics$well_id == well],
                                     median_reads = well_metrics$median_reads[well_metrics$well_id == well],
                                     median_umis = well_metrics$median_umis[well_metrics$well_id == well],
                                     median_genes = well_metrics$median_genes[well_metrics$well_id == well])
}

knitr::kable(well_metrics,
             caption = "Well Metrics Overview")
```

#### Reads per Well Plot
```{r}
qc_violin_plot(all_meta,
               category_x = "well_id",
               name_x = "Well ID",
               column_y = "n_reads",
               name_y = "N Reads per Cell",
               fill = "dodgerblue") +
  ggtitle("Reads per Cell")

```

#### UMIs per Well Plot
```{r}
qc_violin_plot(all_meta,
               category_x = "well_id",
               name_x = "Well ID",
               column_y = "n_umis",
               name_y = "N UMIs per Cell",
               fill = "purple") +
  ggtitle("UMIs per Cell")

```

#### Genes per Well Plot
```{r}
qc_violin_plot(all_meta,
               category_x = "well_id",
               name_x = "Well ID",
               column_y = "n_genes",
               name_y = "N Genes per Cell",
               fill = "orangered") +
  ggtitle("Genes per Cell")

```

[Return to Contents](#contents)

<a id="well_category_counts"></a>

### HTO Category Distribution per Well

#### HTO Category Counts Table
```{r HTO Category Count By Well}
hto_category_by_well <- all_meta[, 
                                 .(n_cells = nrow(.SD)),
                                 by = list(well_id, hto_category)]

hto_category_count_by_well <- dcast(hto_category_by_well,
                                    formula = hto_category ~ well_id,
                                    value.var = "n_cells")

for(well in names(hto_category_count_by_well)[-1]) {
  hto_cat_list <- as.list(hto_category_count_by_well[[well]])
  names(hto_cat_list) <- paste0("n_",hto_category_count_by_well$hto_category)
  qc_list$well_stats[[well]] <- c(qc_list$well_stats[[well]], hto_cat_list)
}

qc_table(hto_category_count_by_well)
```

#### HTO Category Counts per Well Plot
```{r HTO Category Count By Well Plot, fig.height = 6, fig.width = 10}
qc_aligned_barplot(all_meta,
                   category_x = "well_id",
                   name_x = "Well ID",
                   category_y = "hto_category",
                   category_name = "HTO Category",
                   colorset_y = "varibow",
                   name_y = "N Cells",
                   padding = 0.2) +
  ggtitle("HTO Category Counts per Well")
```

#### HTO Category Fraction per Well Table
```{r HTO Category Fraction By Well}
hto_category_by_well <- hto_category_by_well[,
                                             frac_cells := round(n_cells/sum(n_cells),4),
                                             by = well_id]

hto_category_frac_by_well <- dcast(hto_category_by_well,
                                   formula = hto_category ~ well_id,
                                   value.var = "frac_cells")

qc_table(hto_category_frac_by_well)
```

#### HTO Category Fraction per Well Plot
```{r HTO Category Fraction By Well Plot, fig.height = 6, fig.width = 10}
qc_stacked_barplot(all_meta,
                   category_x = "well_id",
                   name_x = "Well ID",
                   category_y = "hto_category",
                   category_name = "HTO Category",
                   colorset_y = "varibow",
                   name_y = "Fraction of Cells",
                   as_fraction = TRUE) +
  ggtitle("HTO Category Fraction per Well")
```

[Return to Contents](#contents)

<a id="well_barcode_counts"></a>

### PBMC Sample/HTO Barcode Distribution per Well

#### HTO Barcode Counts per Well
```{r HTO Barcode Count By Well}
hto_barcode_by_well <- singlet_meta[, 
                                    .(n_cells = nrow(.SD)),
                                    by = list(well_id, hto_barcode)]

hto_barcode_count_by_well <- dcast(hto_barcode_by_well,
                                   formula = hto_barcode ~ well_id,
                                   value.var = "n_cells")

hto_barcode_count_by_well <- hto_barcode_to_pbmc_id[hto_barcode_count_by_well,
                                                    on = "hto_barcode"]

for(well in names(hto_category_count_by_well)[-1]) {
  pbmc_sample_list <- as.list(hto_barcode_count_by_well[[well]])
  names(pbmc_sample_list) <- hto_barcode_count_by_well$pbmc_sample_id
  qc_list$well_stats[[well]]$pbmc_sample_singlet_counts <- pbmc_sample_list
}

for(sample_id in hto_barcode_count_by_well$pbmc_sample_id) {
  sample_row <- hto_barcode_count_by_well[hto_barcode_count_by_well$pbmc_sample_id == sample_id, ,drop = FALSE] 
  sample_row <- sample_row[,-c(1,2)]
  
  qc_list$pbmc_sample_stats[[sample_id]]$well_counts <- as.list(sample_row)
}

qc_table(hto_barcode_count_by_well)
```

#### HTO Barcode Counts per Well Plot
```{r HTO Barcode Count By Well Plot, fig.height = 6, fig.width = 10}
qc_aligned_barplot(singlet_meta,
                   category_x = "well_id",
                   name_x = "Well ID",
                   category_y = "plot_barcode",
                   category_name = "HTO Barcode",
                   colorset_y = "varibow",
                   name_y = "N Cells",
                   padding = 0.2) +
  ggtitle("HTO Barcode Distribution in Wells")
```

#### Well Counts per HTO Barcode Plot
```{r Wells by HTO Barcode Count Plot, fig.height = 6, fig.width = 10}
qc_aligned_barplot(singlet_meta,
                   category_x = "plot_barcode",
                   name_x = "HTO Barcode",
                   category_y = "well_id",
                   category_name = "Well ID",
                   colorset_y = "varibow",
                   name_y = "N Cells",
                   padding = 0.2) +
  ggtitle("Well Distribution in HTO Barcodes")
```

#### HTO Barcode Fraction per Well
```{r HTO Barcode Fraction By Well}
hto_barcode_by_well <- hto_barcode_by_well[,
                                           frac_cells := round(n_cells/sum(n_cells),4),
                                           by = well_id]

hto_barcode_frac_by_well <- dcast(hto_barcode_by_well,
                                  formula = hto_barcode ~ well_id,
                                  value.var = "frac_cells")

hto_barcode_frac_by_well <- hto_barcode_to_pbmc_id[hto_barcode_frac_by_well,
                                                   on = "hto_barcode"]

qc_table(hto_barcode_frac_by_well)
```

#### HTO Barcode Fraction per Well Plot
```{r HTO Barcode Fraction By Well Plot, fig.height = 6, fig.width = 10}
qc_stacked_barplot(singlet_meta,
                   category_x = "well_id",
                   name_x = "Well ID",
                   category_y = "plot_barcode",
                   category_name = "HTO Barcode",
                   colorset_y = "varibow",
                   name_y = "Fraction of Cells",
                   as_fraction = TRUE) +
  ggtitle("HTO Barcode Fraction per Well")
```

#### Well Fraction per HTO Barcode Plot
```{r Well Fraction By HTO Barcode Plot, fig.height = 6, fig.width = 10}
qc_stacked_barplot(singlet_meta,
                   category_x = "plot_barcode",
                   category_y = "well_id",
                   category_name = "Well ID",
                   name_x = "HTO Barcode",
                   colorset_y = "varibow",
                   name_y = "Fraction of Cells",
                   as_fraction = TRUE) +
  ggtitle("Well Fraction per HTO Barcode")
```

[Return to Contents](#contents)

<a id="hash_qc"></a>

### Sample/HTO Barcode QC

<a id="hash_counts"></a>

#### Singlet Counts, Median Reads, UMIs, Genes per HTO Barcode
```{r}
hto_barcode_metrics <- singlet_meta[,.(n_cells = nrow(.SD),
                                       median_reads = as.numeric(median(n_reads)),
                                       median_umis = as.numeric(median(n_umis)),
                                       median_genes = as.numeric(median(n_genes))),
                                    by = hto_barcode]

hto_barcode_metrics <- hto_barcode_to_pbmc_id[hto_barcode_metrics,
                                              on = "hto_barcode"]

for(sample_id in hto_barcode_metrics$pbmc_sample_id) {
  sample_row <- hto_barcode_metrics[hto_barcode_metrics$pbmc_sample_id == sample_id, , drop = FALSE]
  qc_list$pbmc_sample_stats[[sample_id]] <- as.list(sample_row[, -2])
}

qc_table(hto_barcode_metrics)
```

[Return to Contents](#contents)

<a id="hash_reads"></a>

#### Reads per HTO Category/Barcode
```{r}
category_reads_violins <- qc_violin_plot(all_meta,
                                         category_x = "hto_category",
                                         name_x = "HTO Category",
                                         column_y = "n_reads",
                                         name_y = "N Reads per Cell",
                                         fill = "dodgerblue")

barcode_reads_violins <- qc_violin_plot(all_meta[all_meta$hto_category == "singlet",],
                                        category_x = "plot_barcode",
                                        name_x = "HTO Barcode (singlets)",
                                        column_y = "n_reads",
                                        name_y = "N Reads per Cell",
                                        fill = "dodgerblue") +
  ggtitle("Reads per Cell")

reads_violin_list <- list(category_reads_violins, 
                          barcode_reads_violins)

plot_grid(plotlist = reads_violin_list,
          ncol = 2, rel_widths = c(1, 3),
          nrow = 1, align = "h")
```


#### Median Reads per Well By HTO Barcode
```{r}
reads_hto_barcode_by_well <- singlet_meta[,
                                          .(median_reads = median(n_reads)),
                                          by = list(well_id, hto_barcode)]

reads_hto_barcode_by_well <- dcast(reads_hto_barcode_by_well,
                                   formula = hto_barcode ~ well_id,
                                   value.var = "median_reads")

reads_hto_barcode_by_well <- hto_barcode_to_pbmc_id[reads_hto_barcode_by_well,
                                                    on = "hto_barcode"]

for(well in names(reads_hto_barcode_by_well)[-c(1,2)]) {
  pbmc_sample_list <- as.list(reads_hto_barcode_by_well[[well]])
  names(pbmc_sample_list) <- reads_hto_barcode_by_well$pbmc_sample_id
  
  qc_list$well_stats[[well]]$pbmc_sample_mapped_reads_median <- pbmc_sample_list
}

for(sample_id in reads_hto_barcode_by_well$pbmc_sample_id) {
  sample_row <- reads_hto_barcode_by_well[reads_hto_barcode_by_well$pbmc_sample_id == sample_id, ,drop = FALSE] 
  sample_row <- sample_row[,-c(1,2)]
  
  qc_list$pbmc_sample_stats[[sample_id]]$well_mapped_reads_median <- as.list(sample_row)
}

qc_table(reads_hto_barcode_by_well)
```

[Return to Contents](#contents)

<a id="hash_umis"></a>

#### UMIs per HTO Category/Barcode
```{r}
category_umis_violins <- qc_violin_plot(all_meta,
                                        category_x = "hto_category",
                                        name_x = "HTO Category",
                                        column_y = "n_umis",
                                        name_y = "N UMIs per Cell",
                                        fill = "purple")

barcode_umis_violins <- qc_violin_plot(all_meta[all_meta$hto_category == "singlet",],
                                       category_x = "plot_barcode",
                                       name_x = "HTO Barcode (singlets)",
                                       column_y = "n_umis",
                                       name_y = "N UMIs per Cell",
                                       fill = "purple") +
  ggtitle("UMIs per Cell")

umis_violin_list <- list(category_umis_violins, 
                         barcode_umis_violins)

plot_grid(plotlist = umis_violin_list,
          ncol = 2, rel_widths = c(1, 3),
          nrow = 1, align = "h")
```

#### Median UMIs per Well By HTO Barcode
```{r}
umis_hto_barcode_by_well <- singlet_meta[,
                                         .(median_umis = median(n_umis)),
                                         by = list(well_id, hto_barcode)]

umis_hto_barcode_by_well <- dcast(umis_hto_barcode_by_well,
                                  formula = hto_barcode ~ well_id,
                                  value.var = "median_umis")

umis_hto_barcode_by_well <- hto_barcode_to_pbmc_id[umis_hto_barcode_by_well,
                                                   on = "hto_barcode"]

for(well in names(umis_hto_barcode_by_well)[-c(1,2)]) {
  pbmc_sample_list <- as.list(umis_hto_barcode_by_well[[well]])
  names(pbmc_sample_list) <- umis_hto_barcode_by_well$pbmc_sample_id
  
  qc_list$well_stats[[well]]$pbmc_sample_umis_median <- pbmc_sample_list
}

for(sample_id in umis_hto_barcode_by_well$pbmc_sample_id) {
  sample_row <- umis_hto_barcode_by_well[umis_hto_barcode_by_well$pbmc_sample_id == sample_id, ,drop = FALSE] 
  sample_row <- sample_row[,-c(1,2)]
  
  qc_list$pbmc_sample_stats[[sample_id]]$well_umis_median <- as.list(sample_row)
}

qc_table(umis_hto_barcode_by_well)
```

[Return to Contents](#contents)

<a id="hash_genes"></a>

#### Genes per HTO Category/Barcode
```{r}
category_genes_violins <- qc_violin_plot(all_meta,
                                         category_x = "hto_category",
                                         name_x = "HTO Category",
                                         column_y = "n_genes",
                                         name_y = "N Genes per Cell",
                                         fill = "orangered")

barcode_genes_violins <- qc_violin_plot(all_meta[all_meta$hto_category == "singlet",],
                                        category_x = "plot_barcode",
                                        name_x = "HTO Barcode (singlets)",
                                        column_y = "n_genes",
                                        name_y = "N Genes per Cell",
                                        fill = "orangered") +
  ggtitle("Genes per Cell")

genes_violin_list <- list(category_genes_violins, 
                          barcode_genes_violins)

plot_grid(plotlist = genes_violin_list,
          ncol = 2, rel_widths = c(1, 3),
          nrow = 1, align = "h")
```

#### Median Genes per Well By HTO Barcode
```{r}
genes_hto_barcode_by_well <- singlet_meta[,
                                          .(median_genes = median(n_genes)),
                                          by = list(well_id, hto_barcode)]

genes_hto_barcode_by_well <- dcast(genes_hto_barcode_by_well,
                                   formula = hto_barcode ~ well_id,
                                   value.var = "median_genes")

genes_hto_barcode_by_well <- hto_barcode_to_pbmc_id[genes_hto_barcode_by_well,
                                                    on = "hto_barcode"]

for(well in names(genes_hto_barcode_by_well)[-c(1,2)]) {
  pbmc_sample_list <- as.list(genes_hto_barcode_by_well[[well]])
  names(pbmc_sample_list) <- genes_hto_barcode_by_well$pbmc_sample_id
  
  qc_list$well_stats[[well]]$pbmc_sample_genes_median <- pbmc_sample_list
}

for(sample_id in genes_hto_barcode_by_well$pbmc_sample_id) {
  sample_row <- genes_hto_barcode_by_well[genes_hto_barcode_by_well$pbmc_sample_id == sample_id, ,drop = FALSE] 
  sample_row <- sample_row[,-c(1,2)]
  
  qc_list$pbmc_sample_stats[[sample_id]]$well_genes_median <- as.list(sample_row)
}

qc_table(genes_hto_barcode_by_well)
```

[Return to Contents](#contents)

<a id="json_out"></a>

### Write QC JSON

```{r Save QC JSON}
stm(paste0("Writing JSON to ", out_json))

qc_list_json <- jsonlite::toJSON(qc_list,
                                 auto_unbox = TRUE,
                                 pretty = TRUE)

writeLines(qc_list_json,
           out_json)
```

[Return to Contents](#contents)

<a id="session_info"></a>

## Session Information

```{r Session Info}
sessionInfo()
```

Total time elapsed
```{r Show Time}
end_time <- Sys.time()
diff_time <- end_time - start_time
time_message <- paste0("Elapsed Time: ", 
                       round(diff_time, 3),
                       " ", units(diff_time))
print(time_message)
stm(time_message)
stm("H5 merge process complete.")
```

[Return to Contents](#contents)
